working owls methods

### 2020-11-18 manuscript_working/ms_working_RMD.Rmd
updating manuscript wrt MRE edits

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/11/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

*Table 1XX. Species detection records by route. Ones indicate that that species was detected during that route and year, and zeros indicated that those owls were not detected, although they may have been present and undetected. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador, although specific survey years varied by route, which was given in column "Year"*

``` {r}
load("../data/output_data/richness_species_accounts_byRt.Rdata")
library(pander)
pander(species.accounts.byRt)
```

## Occupancy Model Framework

Let

\[
\begin{aligned}
t  &=  \text{the years of surveys (here }t = 1,2,\ldots,11)\text{ corresponding with 2003-2013}\\
h  &=  \text{the individual routes (here }h = 1,2,\ldots,6)\text{ corresponding with EI1, EI2, M1, M2, N1, N2}\\
i  &=  \text{the individual surveys conducted in each year and route (here }i = 1,2,3)\\
j  &=  \text{the individual stations along each route (here }j = 1,2,\ldots,10)\\
k  &=  \text{the broadcast species (here }k = 1,2,\ldots,10)\\
z_{t,h,i} &= \text{a random variable equal to 1 when a survey was occupied and 0 otherwise}\\
y_{t,h,i,j,k} &= \text{a random variable equal to 1 when an owl was detected and 0 otherwise}
\end{aligned}
\]

We assumed that survey occupancy ($z_{t,h,i}$) was the outcome of Bernoulli trials with probability $\psi$, which we allowed to vary by year and route:

@@

We assumed that detecting an owl depended on an owl being present during that survey ($z_{t,h,i} = 1$) and the probability of detection, which was related to the species of owl played for broadcast ($k$):

@@

This model provides a consistent probability of detection for all surveys in the first two minutes of each survey before the broadcast owl recording was played ($k = \text{Pre-broadcast}$). Then, the probability of detection for each post-broadcast time period would depend on the species of broadcast owl. This allowed species-specific behavior in response to the different broadcast owl species [@Baumgardt:2019]. We used means parameterization such that the coefficients were interpretable as the effect of that specific broadcast species $k$ (including the pre-broadcast time frame). The priors for every logistic model coefficient $\beta_k$ were chosen to be near uniform as recommended in @Gelman:2008. Specifically, we used the Cauchy prior such that: $\beta_k \sim \text{Cauchy}(\text{precision} = 0.16)$.

The broadcast species were consistent at each station across years, but varied by route (Table 2XX).  Thus, the effective sample sizes of the different broadcast species varied.

*Table 2XX. Broadcast species by route and station. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Pre-recorded owl calls were broadcast after 2 minutes of silent listening.*

@@

### Occupancy Model Implementation

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the occupancy model for three owl species: Mottled, Spectacled, and FerPy owls. These owls had enough positive detections to analyze occupancy, as Mottled, Spectacled, and FerPy owls had 542, 137, and 187 positive detections over the 11 year period, respectively. Based on our understanding of owl ecology, we assumed that Spectacled and FerPy owls would not occupy route M1 in Montecristo, so we removed M1 from those species' occupancy analyses.

@@

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 10 species observed during our surveys (including unknown owl species) and an additional 4 unidentified species. This limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus one species we identified (Stygian owl), which was previously undocumented in El Salvador. We augmented our owl detection data with 4 additional potential species with all-zero detection [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,14$), $w_{t,h,s}$ was a binary indicator of that owl species being present in each route and year:

$$
w_{t,h,s} \sim \text{Bernoulli}(\Omega_h)
$$
where $\Omega_h$ was the probability of species presence in each route $h$. In the richness model, the presence of an owl species in any route, year, and survey, $z_{t,h,i,s}$, was conditional on the probability of occupancy for that route, year, and species $\psi_{t,h,s}$ *and* that species belonging to that route's community ($w_{t,h,s}$):

@@

Richness was conditional on each species presence and was estimated as the sum of owl species present in each route and year:

@@

\[
\begin{aligned}
\psi_{t,h,s} &\sim \text{Beta}(a^\psi_{t,h,s},b^\psi_{t,h,s})\\ 
a^\psi_{t,h,s} &= \mu^\psi_{t,h}\rho^\psi\\
b^\psi_{t,h,s} &= \rho^\psi - \mu^\psi_{t,h}\rho^\psi\\
\text{logit}(\mu^\psi_{t,h}) &= \text{E}(\psi_{t,h,s}) \sim N(0,2.25)\\
\ln(\rho^\psi) &\sim N(5,1)\text{, truncated to (0.01,10)}\\
\end{aligned}
\]

Consistent with our approach for the occupancy models, we assumed that detecting an owl depended on an owl of that species being present during that survey ($z_{t,h,i,s} = 1$) and the probability of detection $p_{t,h,i,j,k,s}$, which was related to the broadcast species:

@@

where generally $\text{logit}(p_{t,h,i,j,k,s}) = \beta_kX_k$ for each broadcast species $k$ as described above. The coefficients of the detection model were thus assumed to be consistent for all owl species, which allowed us to borrow information about the detection process for undetected owls from those that were detected.

@@

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the richness model. 
We ran 3 chains for 20,000 iterations and 2000 iterations discarded as burn-in and a thinning rate of 2, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i,s}$ were equivalent to one if that species of owl $s$ was detected in that survey $i$ and otherwise zero. Similarly, initial values for $w_{t,h,s}$ were equivalent to one if that species of owl was detected in that route and year and otherwise zero. We visually inspected traceplots to verify that chains mixed well (Supplemental information S4xx).

@@

## Occupancy Model Results

Averaged across years, we found that the probability of occupancy was somewhat higher for Mottled owls than for FerPy or Spectacled owls in both routes of El Imposible and in M2 route of Montecristo (Fig. 1X). The probability of occupancy across all three species and both routes of Nancuchiname were relatively high (Fig. 1XX). We found that the probability of occupancy for Mottled owls was relatively consistent across time in El Imposible and in N2 route (Fig. 2XX), but that it seemed to increase in the last 6 or so years of the surveys in M2 and N1 routes (Fig. 2XX). The probability of occupancy in M1 seemed to have decreased through time for Mottled owls (Fig. 2XX).

The probability of occupancy for FerPy and Spectacled stayed relatively consistently low across time in routes EI1, EI2, and M2 other than a higher probability of occupancy for FerPy in EI1 in the early years of the surveys (Figs. 3XX and 4XX). FerPy probability of occupancy was relatively high in both routes of Nancuchaname except in 2008 and 2009 (Fig. 3XX), and the probability of occupancy for the Spectacled owl slightly fluctuated year to year in the Nancuchaname routes (Fig. 4XX).

@@

Adding in all detected species to one richness model increased the precision of the estimates of the effect of broadcast species on detection probability (i.e., compare All Species with individual occupancy results in Fig. 6XXX). When incorporating all owl species, broadcasting calls from Mottled, Pacific, Crested, Black and White, and Spectacled Owls increased detection probability; whereas broadcasting the Great Horned owl calls decreased detection (Fig. 6XXX).

The probability of detection when predicted through independent occupancy analyses varied little during the timeframe before broadcast recordings were played (Pre-broadcast); however, the probability of detecting owls after broadcast species were called varied by broadcast species and between Mottled, FerPy, and Spectacled owls (Fig. 6XX). The probability of detecting Mottled Owls increased after broadcasting Mottled, Pacific, Crested, Black and White, or Spectacled owls. The probability of detecting FerPy owls increased after broadcasting Mottled and Pacific owl calls, and the probability of detecting Spectacled owls increased after broadcasting Pacific, Black and White, and Spectacled owls (Fig. 6XX).

@@

![1XX. Probability of occupancy ($\psi_{t,h}$) averaged by route for Mottled, FerPy, and Spectacled Owls. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. \label{psi_means}](../output/figures/psi_means-1.png)

@@

![3XX. Probability of occupancy by route and year for FerPy Owl. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. \label{figure3}](../output/figures/ferpy_psi_byYr-1.png)

>>>>

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/17/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

*Table 1XX. Species detection records by route. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador, although specific survey years varied by route.*

``` {r}
load("../data/output_data/richness_species_accounts_byRt.Rdata")
library(pander)
pander(species.counts.long[c(3:9)])
```

## Single-Species Occupancy Model Framework

We modeled occupancy within each route, year, and survey assuming that the probability of occupancy would be closed across surveys of a given route in a given year. In other words, for any single species of owl, we assumed that the probability of a route being occupied by that owl species would not change between surveys within a year but that the probability of occupancy would vary between years. Let:

\[
\begin{aligned}
t  &=  \text{the year each survey was conducted (here }t = 1,2,\ldots,11)\text{ corresponding with 2003-2013}\\
h  &=  \text{the individual route (here }h = 1,2,\ldots,6)\text{ corresponding with EI1, EI2, M1, M2, N1, N2}\\
i  &=  \text{the individual surveys conducted in each year and route (here }i = 1,2,3)\\
j  &=  \text{the individual stations along each route (here }j = 1,2,\ldots,10)\\
k  &=  \text{the broadcast call species (here }k = 1,2,\ldots,10)\\
z_{t,h,i} &= \text{a random variable equal to 1 when a survey was occupied and 0 otherwise}\\
y_{t,h,i,j,k} &= \text{a random variable equal to 1 when an owl was detected and 0 otherwise}
\end{aligned}
\]

We assumed that occupancy in each individual survey ($z_{t,h,i}$) was the outcome of Bernoulli trials with a probability of occupancy, $\psi_{t,h}$, which we allowed to vary by year and route:

@@

We assumed that detecting an owl depended on an owl being present during that survey ($z_{t,h,i} = 1$) and the probability of detection, which was related to the species-specific broadcast call ($k$):

@@

This model provides a consistent probability of detection for all surveys in the first two minutes of each observation period, before the broadcast call was played ($k = \text{Pre-broadcast}$). Then, the probability of detection for each post-broadcast time period depended which call had been played. This allowed species-specific behavior in response to the different broadcast calls [@Baumgardt:2019]. We used means parameterization for the logistic regression model such that the coefficients were interpretable as the effect of that specific broadcast call $k$ (including the pre-broadcast time frame). The priors for every logistic model coefficient $\beta_k$ were chosen to be near uniform as recommended in @Gelman:2008: $\beta_k \sim \text{Cauchy}(\text{precision} = 0.16)$.

The broadcast calls varied between routes, but were consistent at each route and station across years (Table 2XX).  Thus, the effective sample sizes of the different broadcast calls varied.

*Table 2XX. Broadcast calls by route and station. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Pre-recorded owl calls were broadcast after 2 minutes of silent listening.*

@@

### Single-Species Occupancy Model Implementation

We used the R2jags package in R [@R:2014] [@Plummer:2013] to implement the occupancy model for three owl species: Mottled, Spectacled, and Ferruginous Pygmy owls. These species had sufficient positive detections to analyze occupancy (Table 1XX). Based on our understanding of owl ecology, we assumed that Spectacled and Ferruginous Pygmy owls would not occupy route M1 in Montecristo, so we removed M1 from those species' occupancy analyses.

@@

To model species richness at each route in each year, we assumed that there were 15 possible species present in El Salvador. This included the 10 species observed during our surveys (including unknown owl species) and an additional 5 species that we never observed during our surveys (Table 1XX). This upper limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus our unknown species and one species we identified (Stygian owl), which was previously undocumented in El Salvador. We thus augmented our owl detection data with 5 additional potential species with all-zero detection records [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,15$), $w_{t,h,s}$ was a binary indicator of whether that owl species was present in each route and year:

$$
w_{t,h,s} \sim \text{Bernoulli}(\Omega_h)
$$
where $\Omega_h$ was the probability of species presence in each route $h$ [@Broms:2016]. We used prior recommended by @Broms:2016, @Guillera-Arroita:2019, and others for $\Omega_h$: $\Omega_h \sim \text{Beta}(0.001, 1)$.

In the richness model, the presence of an owl species in any route, year, and survey, $z_{t,h,i,s}$, was conditional on the probability of occupancy for that route, year, and species $\psi_{t,h,s}$ *and* that species belonging to that route's community that year ($w_{t,h,s}$):

@@

\[
\begin{aligned}
\psi_{t,h,s} &\sim \text{Beta}(a^\psi_{t,h,s},b^\psi_{t,h,s})\\ 
a^\psi_{t,h,s} &= \mu^\psi_{t,h}\rho^\psi\\
b^\psi_{t,h,s} &= \rho^\psi - \mu^\psi_{t,h}\rho^\psi\\
\text{logit}(\mu^\psi_{t,h}) &= \sim N(0,2.25)\\
\ln(\rho^\psi) &\sim N(5,1)\text{, truncated to (0.01,10)}\\
\end{aligned}
\]

Richness was conditional on each species' presence and was estimated as the sum of owl species present in each route and year:

@@

Consistent with our approach for the occupancy models, we assumed that detecting an owl depended on an individual of that species being present during that survey ($z_{t,h,i,s} = 1$) and the species-specific probability of detection $p_{t,h,i,j,k,s}$, which was related to the broadcast call:

@@

where generally $\text{logit}(p_{t,h,i,j,k,s}) = \beta_kX_k$ for each broadcast call $k$ as described above. The coefficients of the detection model were thus assumed to be consistent for all owl species, which allowed us to borrow information about the detection process for undetected owl species from those that were detected.

@@

We used the R2jags package in R [@R:2014] [@Plummer:2013] to implement the richness model. We ran 3 chains for 20,000 iterations with 2000 iterations discarded as burn-in and a thinning rate of 2, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i,s}$ were equivalent to one if that species of owl was detected in that survey and otherwise zero. Similarly, initial values for whether a species belonged to a route's community ($w_{t,h,s}$) were equivalent to one if that species of owl was detected in that route and year and otherwise zero. We visually inspected traceplots to verify that chains mixed well (Supplemental information S4xx).

@@

## Single-Species Occupancy Model Results

Mottled, Spectacled, and Ferruginous Pygmy owls had 542, 137, and 187 positive detections over the 11 year period, respectively (Table 1XX). No other owl species had enough detections to analyze species-specific occupancy (Table 1XX). Averaged across years, we found that the probability of occupancy was somewhat higher for Mottled owls than for Ferruginous Pygmy or Spectacled owls in both routes of El Imposible and in M2 route of Montecristo (Fig. 1X). The probability of occupancy across all three species and both routes of Nancuchiname were relatively high (Fig. 1XX). We found that the probability of occupancy for Mottled owls was relatively consistent across time in El Imposible and in N2 route (Fig. 2XX), but that it seemed to increase in the last ~6 years of the surveys in M2 and N1 routes (Fig. 2XX). The probability of occupancy in M1 seemed to have decreased through time for Mottled owls (Fig. 2XX).

For routes EI1, EI2, and M2, the probability of occupancy for Ferruginous Pygmy and Spectacled stayed relatively low across time other than a higher probability of occupancy for Ferruginous Pygmy in EI1 in the early years of the surveys (Figs. 3XX and 4XX). The probability of occupancy for Ferruginous Pygmy owls was relatively high in both routes of Nancuchaname except in 2008 and 2009 (Fig. 3XX), and the probability of occupancy for the Spectacled owl slightly fluctuated year to year in the Nancuchaname routes (Fig. 4XX).

@@

Including all detected species into one richness model increased the precision of the estimates of the effect of broadcast calls on detection probability (i.e., compare All Species with individual occupancy results in Fig. 6XXX). When incorporating all owl species, broadcasting calls from Mottled, Pacific, Crested, Black and White, and Spectacled Owls increased detection probability; whereas broadcasting the Great Horned owl calls decreased detection (Fig. 6XXX).

The probability of detection when analyzed with single-species occupancy models varied little during the timeframe before broadcast recordings were played (Pre-broadcast); however, the probability of detecting owls after broadcast calls were played varied by broadcast call and between Mottled, Ferruginous Pygmy, and Spectacled owls (Fig. 6XX). The probability of detecting Mottled Owls increased after broadcasting Mottled, Pacific, Crested, Black and White, or Spectacled owl calls. The probability of detecting Ferruginous Pygmy owls increased after broadcasting Mottled and Pacific owl calls, and the probability of detecting Spectacled owls increased after broadcasting Pacific, Black and White, and Spectacled owl calls (Fig. 6XX).

@@

![1XX. Probability of occupancy ($\psi_{t,h}$) averaged by route for Mottled, Ferruginous Pygmy, and Spectacled Owls. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. \label{psi_means}](../output/figures/psi_means-1.png)

@@

![3XX. Probability of occupancy by route and year for Ferruginous Pygmy Owl. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. \label{figure3}](../output/figures/ferpy_psi_byYr-1.png)
